---
title: "Tutorial"
author: "Mandy Liesch"
output: html_document
code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


```

# Packages

## Data Downloading

`geodata` : The [geodata](https://github.com/rspatial/geodata) package has functions for downloading of geographic data for use in spatial analysis and mapping. The package facilitates access to climate, crops, elevation, land use, soil, species occurrence, accessibility, administrative boundaries and other data.

`tigris` : The [tigris](https://github.com/walkerke/tigris) package allows for the downloading of spatial data at several geographic scales, including states and counties (shown here).

## Geoprocessing

`terra` : The [terra](https://rspatial.org/pkg/#google_vignette) package allows for methods for spatial data analysis with vector (points, lines, polygons) and raster (grid) data. This package is faster than the raster package. 

`tidyterra` : The [tidyterra](https://dieghernan.github.io/tidyterra/reference/ggspatvector.html) package provides functions designed to seamlessly integrate terra objects (like SpatVector) with ggplot2. 

## Display

`ggpubr` the [ggpubr](https://rpkgs.datanovia.com/ggpubr/) package provides some easy-to-use functions for creating and customizing 'ggplot2'- based plots creating publication ready plots.

`gridExtra` the [gridExtra](https://cran.r-project.org/web/packages/gridExtra/vignettes/arrangeGrob.html) package allows for the display of plots side by side. 

`tidyverse` the [tidyverse](https://www.tidyverse.org/) package is an opinionated collection of R packages designed for data science. All packages share an underlying design philosophy, grammar, and data structures.



```{r, include=FALSE}
library(tigris)
library(tidyverse)
library(terra)
library(tidyterra)
library(gridExtra)
library(ggpubr)
library(geodata)


```

# Vector Data

Vector Data is commonly used at its original resolution, like most traditional maps historically. This type of data comes in lines, and dots (or vertexes). These two elements are put together to create polygons, like a boundary of a lake. It can maintain data and create smooth network analysis. The vertexes need to be stored explicitly in the data, and sometimes, can require a lot of data cleaning. Whenever a file is changed, the underlying network (topology) has to be built again. In this case, we are working with polygon vector data that are the county data shapefiles. 

## Shapefiles

There are 3233 county polygons and 9 data fields in the intial tigris file. However, the state vector (using State FIPS), narrows the states to ones that are in the Mississippi/Ohio River Basin. We are looking at a 12 state area that make up the bulk of these basins.

```{r echo=T, results='hide'}
#Set the option to remove scientific notation from displays
options(scipen=10000)

#Create a vector with the state FIPS that we need to extract. 
states<-c("5", "17", "18", "19", "21", "22", "27","28",  "29", "39", "47", "55")

#Use the tigris package, and get the counties extracted by the state we specificied
#in the states vector
counties_raw <- tigris::counties(states,
  #add the county boundaries
  cb = TRUE,
  #specify the resolution (we don't need the highest one)
  resolution = "500k",
  #Pick the year the county files are extracted
  year = 2019,
  #specify the data type
  class = "sf"
)
```

```{r}
#Check the metadata information
counties_raw 

```

The polygon file has several important attributes, as well as columns of data associated with each county file (shapefile). There are 1091 counties, with 9 different geographic fields. 

Bounding box limits the geographic area. The Geodetic CRS is very important information when we are dealing with spatial data. IF you are combining or layering information, this value must be the same in both sources, or projections will not line up. The NAD83 is very common for US Datasets.  

Columns include the State FIP value and the county FIP values, as well as the combined GEOID, County Name, Area of Land, water and the Geometry attributes. We do not need the extra values, so we can remove all of the extra columns, except for GEOID and the county NAME. 

```{r, results = FALSE}
#Select the columns needed for geographic data
counties <- counties_raw %>%  
  #Choosing the GEOID to pair with datasets and the county Names
  select(GEOID, NAME)

#Convert data file into a spatVector
counties<- vect(counties)

#plot the counties to check the geographic area
plot(counties)

```

When you use the plot function in base R, it will plot every column as a unique plot, in this case, we will get 2 of these plots. Changing the values into a spatvector using the Terra package shows only one shapefile, and removes the geometries as a column. This also allows the setup to use values in ggplot at a later time. 


## Ancillary Data

Most of these shapefiles can be merged with datafiles and spreadsheets to expand and visualize anything with data. A majority of this federally based data is classified by State code (two numbers) and county code (three numbers). These are combined together to create a GEOID (or FIPS).  For example, the State of Wisconsin has a state code of 55, and each county has its own number, Outagamie County, WI, for example, is county 087, giving it an idenity of 55087. 

Almost all federal agencies have large respositories of data that anyone can access, via downloads of spreadsheet files (or .csv) that can be used for mapping with shapefiles. 

[Atlas of Rural Towns (USDA ERS)](https://www.ers.usda.gov/data-products/atlas-of-rural-and-small-town-america): Focused on statistical metrics for Rural America, including People, Jobs, Classification, Incomes and Veteran based data. 

[USDA ERS Data](https://www.ers.usda.gov/data-products):  ERS has a huge amount of data covering key information about the farm and food sectors, including agricultural markets and trade, farming and farm income, food prices and consumption, food security and nutrition assistance, rural economies, and the interaction of agriculture and natural resources.

[CDC Wonder](https://wonder.cdc.gov/): It provides access to a wide array of public health data for researchers, public health professionals, and the general public. CDC WONDER offers various online databases containing information on births, deaths, cancer, STDs, TB, and more.

[National Agricultural Statistics Service](https://quickstats.nass.usda.gov/): Conducts hundreds of surveys every year and prepares reports covering virtually every aspect of U.S. agriculture. This includes the Census of Agriculture, published every 5 years. Production and supplies of food and fiber, prices paid and received by farmers, farm labor and wages, farm finances, chemical use, and changes in the demographics of U.S. producers are only a few examples. 

[Census Bureau](https://www.census.gov/data.html): A huge source of data, including the American Community Surveys, and the census taken every 10 years. If you need data on demographics, economics, and all things socio-demographic.

[Bureau of Labor Statistics](https://www.bls.gov/data/):  It is the principal fact-finding agency in the broad field of labor economics and statistics and serves as part of the U.S. Federal Statistical System. BLS collects, calculates, analyzes, and publishes data essential to the public, employers, researchers, and government organizations.



## Datafile

For these shapefiles, we are utilizing a .csv file that was curated from the NASS database from the 2017 Agricultural Census for states on the Mississippi River Basin. 

```{r}
#Load the data frame
NASS2017<-read.csv("NASS2017_Snapshot.csv")

#Make sure that all columns have 5 numbers 
NASS2017$FIPS<-sprintf("%05d", NASS2017$FIPS)

#Check example output
head(NASS2017)

```

There are 1,090 counties in this area, and 27 columns of data involving crop and agricultural production data. 

*FIPS:* Federal code identifying counties 

*Market_Value:* Total Agricultural Market Value in the County in dollars

*Per_Crop_sales:* Percentage of Agricultural Income Generated by Crop Production

*Per_Grain_Mark_Val:*  Percentage of Agricultural Income Generated by grain crops (Corn, soybean, etc.)

*Per_Cotton_Mark_Val:*  Percentage of Agricultural Income Generated by cotton

*Per_Other_Crop_Mark_Val:* Percentage of Agricultural Income Generated by other crops (like oil crops, hops, etc)

*Per_Veg_Sale_Mkt_Val:* Percentage of Agricultural Income Generated by vegetables (carrots, cabbage, etc)

*Per_Fruit_Sales_Mkt_Val:* Percentage of Agricultural Income Generated by fruits (melons, cranberries, blueberries, etc)

*Per_Nursery_Sales_Val:*  Percentage of Agricultural Income Generated by nursery crops

*Per_Cut_Tree_Sales_Val:*  Percentage of Agricultural Income Generated by cut trees

*Per_Ani_Sales_Mark_Val:* Percentage of Agricultural Income Generated by Animal Production

*Per_Poul_Sales_Mark_Val:* Percentage of Agricultural Income Generated by Poultry Production (Turkey, Broilers, Eggs, etc.)

*Per_Milk_Sales_Mrk_Val:* Percentage of Agricultural Income Generated by Milk and Milk products

*Per_Cattle_Sales_Mrk_Val:* Percentage of Agricultural Income Generated by live cattle sales

*Per_Hog_Sales_Mrk_Val:* Percentage of Agricultural Income Generated by hog sales

*Per_Sheep_Sales_Mrk_Val:* Percentage of Agricultural Income Generated by sheep sales

*PER_AQ_SALES:* Percentage of Agricultural Income Generated by aquaculture sales

*PER_EQ_SALES:* Percentage of Agricultural Income Generated by equine sales

*PER_SP_AN_SALES:* Percentage of Agricultural Income Generated by other animal sales

*Per_Cropland_Manure:* Percentage of Agricultural Land with Manure application

*Cattle_Operations:* Number of Cattle Operations in the county

*Milk_Operations:* Number of Milk Operations in the county

*Poul_Total_Oper:* Number of Poultry Operations in the county

*Hog_Oper:* Number of Hog Operations in the county

*TOT_SHEEP_Oper:* Number of sheep Operations in the county

*EQUI_OPER:* Number of horse Operations in the county

*TOT_AQUA_OPER:* Number of Aquaculture Operations in the county

In order to visualize this data, we need to merge this database with the shapefile. 

```{r}
#the counties data and Agricultural Statistics file
ag_data<-merge(counties, NASS2017, 
               #The Counties files is called GEOID
               by.x='GEOID', 
               # The NASS file is merged from the FIPS column
               by.y='FIPS')

#Check the data
ag_data
```

The file is now converted into a Spatvector, in the NAD83 reference system.

### Create a Map

Use the Geom Spatvector function to graph the shapefile. The resulting file should be a blank shapefile. We need to add parameters to the ggplots to actually make meaningful maps. 

```{r}
#cue up a ggplot
ggplot(ag_data) +
  #utilize the geom_spatvector in tidyterra
  geom_spatvector()

```

Add a column to display using the AES fill option

```{r}
crop_plot<-ggplot() +
  #using the dataset, the fill = column can be changed to any of the above columns
  #to display another variable
  geom_spatvector(data=ag_data, 
                  #The fill value is the parameter that AES visualizes
                  aes(fill = Per_Crop_Sales))

#Check the plot
crop_plot

```

Get a new color scheme. [Color schemes](https://ggplot2-book.org/scales-colour.html) can be flexible, and displayed in many different ways. There are a million different color options, but the Viridis Pallette are designed to be perceptually uniform in both colour and when reduced to black and white, and to be perceptible to people with various forms of colour blindness. The data we are looking at today is continuous, not categorical, which is where the c comes from in the scale fill. There are 8 different options in the Viridis pallette. 

```{r}
#Look at the different color options
?scales::pal_viridis

#Create the ggplot
crop_plot<-ggplot() +
  #using the dataset, the fill = column can be changed to any of the above columns
  #to display another variable. I am using the Percent Crop Sales here
  geom_spatvector(data=ag_data, aes(fill = Per_Crop_Sales)) +
  #Change the color scale using the option 
    scale_fill_viridis_c(option = "cividis")

#Check the plot
crop_plot

```

Customize the legend name by adding a name function to the scale fill option. 

```{r}

#Plot the plot with a Legend Name
crop_plot<-ggplot() +
  #using the dataset, the fill = column can be changed to any of the above columns
  #to display another variable
  geom_spatvector(data=ag_data, aes(fill = Per_Crop_Sales)) +
  #Change the color scale using his options
    scale_fill_viridis_c(option = "cividis", 
                         #Customize the Name of the Legend
                         name = "Percent")

#Check the Plot
crop_plot

```

Add a Title and change the theme to center the title and put the legend on the bottom of the plot. There are many different [theme elements](https://ggplot2.tidyverse.org/reference/element.html) available to customize plots. These elements also exist with [legend configurations](https://r-graph-gallery.com/239-custom-layout-legend-ggplot2.html)

```{r}
#Look at the different color options
crop_plot<-ggplot() +
  #using the dataset, the fill = column can be changed to any of the above columns
  #to display another variable
  geom_spatvector(data=ag_data, aes(fill = Per_Crop_Sales)) +
  #Change the color scale using his options
    scale_fill_viridis_c(option = "cividis", 
                         #Customize the Name of the Legend
                         name = "Percent") +
  #Add a title
  ggtitle("Percent Sales from Crops") +
  #Use the ggplot theme function and center it using the element text
  theme(plot.title = element_text(hjust=0.5), 
        #And put the legend in the bottom
        legend.position = "right")

#Check the plot
crop_plot

```


### Your Turn! 

Change the display variable to a different column, and then change the color and name! You can scroll up to get the variable names and definitions, or you can use the names(ag_data) to get a list of all of the variables

```{r}
#Look at the legend names
names(ag_data)

#Look at the different color options
?scales::pal_viridis


plot2<-ggplot() +
  #using the dataset, the fill = column can be changed to any of the above columns
  #to display another variable
  geom_spatvector(data=ag_data, aes(fill = Per_Crop_Sales)) +
  #Change the color scale using this options
    scale_fill_viridis_c(option = "cividis", 
                         #Customize the Name of the Legend
                         name = "Percent")  +
  #Customize the title based on your data
  ggtitle("Percent Sales from Crops") +
  #You can put the legend in the center here, or leave it on the left (delete element text)
  #or right ((hjust =1))
  theme(plot.title = element_text(hjust=0.5), 
        #Feel free to move the legend away from the bottom, or hit ?theme for other options
        legend.position = "right")

plot2

```

### Side by Side Plots 

This using the gridExtra package, we can arrange graphs to be side by side into rows or columns. This data is better suited for columns. So we can use the grid.arrange to plot several plots into the same area. 

```{r}
#Using the grid arrange, we can put the two plots together side by side
grid.arrange(crop_plot, plot2, ncol=2)
```


### Side by Side Plots With the Same Legend

If you have data like we do in this NASS dataset that is mostly in percentage (though, we have some count or dollar data). You can extract the legend and turn it into a ggplot figure

```{r}
# Extract the legend. Returns a gtable
leg <- get_legend(crop_plot)

# Convert to a ggplot and print
as_ggplot(leg)

#Remove the legends from the old plots
crop_plot <- crop_plot + theme(legend.position="none")

#Remove the legends from the old plots
plot2 <- plot2 + theme(legend.position="none")

#Create a file with two different plots and the legend into three columns
grid.arrange(crop_plot, plot2, leg, ncol=3, 
             #Customize the widths to allow 
             widths=c(2.3, 2.3, 0.8))
```

### Vector Calculations

You can utilize the data existing in the data frame to create new columns for display as well! We have the total market value, and can use this to multiply by the percent sales to get total market values in the Poultry Industry, and then divide by total producers to get total farm size too! 

```{r}
#Calculate the amount that poultry operations make per operator
ag_data$Poultry_Per_Capita = (ag_data$MARKET_VALUE * (ag_data$Per_Poul_Sales_Mark_Val/100)) / ag_data$Poul_Total_Oper

Poultry_Per_Capita<-ggplot() +
  #using the dataset, the fill = column can be changed to any of the above columns
  #to display another variable
  geom_spatvector(data=ag_data, aes(fill = Poultry_Per_Capita)) +
  #Change the color scale using this options
    scale_fill_viridis_c(option = "magma", 
                         #Customize the Name of the Legend
                         name = "Dollars $") +
  #Customize the title based on your data
  ggtitle("Per Capita Poultry Operation Income") +
  #You can put the legend in the center here, or leave it on the left (delete element text)
  #or right ((hjust =1))
  theme(plot.title = element_text(hjust=0.5), 
        #Feel free to move the legend away from the bottom, or hit ?theme for other options
        legend.position = "right")

Poultry_Per_Capita

```


# Raster Data

Raster data represents information as a grid of tiny squares called pixels. They are super common for use in spatially explicit datasets. Each cell has a position and color and is stored in a matrix. Each layer of maps show one attribute, and they are ideally suited for modeling and analysis. Data can be discrete (like land use), or continuous (like elevation), and it can handle both types very easily. The cell size determines resolution for the data. Size of rasters are variable, and it is difficult to establish linear relationships (they get very pixellated). They only can represent one value at a time, and the data can be very large over time and areas. Data also needs to be generalized or interpolated, resulting in data losses and outputs. 

## Raster Processing

The geodata package is a very diverse package that allows for the deployment of many different types of earth science data. 

```{r}

# Read the .tif  as a SpatRaster
SG_2020<-soil_world_vsi(var="ocs", depth=30)

```

This data is a global spatraster with  58034 x 159246 cells of a 250 m x 250 m resolution. And the resolution is the Interrupted Goode Homolosine. In order to use the data with our existing agricultural dataset, we have to project the raster into the same coordinate reference system as the agricultural shapefile. We do this using project(from, to). Then we crop and mask the function of the global soil map to the extent of the shapefile. This makes sure the area of the global map and clip it to the ag_data1 file. 

```{r echo=T, results='hide'}

#Project the ag_data from the NAD 83 into the SG_2020 IGH Dataset project(from, to)
ag_data1<-project(ag_data, SG_2020)

#The crop function cuts the FAO glosis map to the ag_data1 area. Mask =  True cuts away the excess. 

MS_Area<-crop(SG_2020, ag_data1, 
              #Remove this function and see what it does! 
              mask=TRUE)

MS_Area

```

When we check the projection, we can see the area is down to 9007 x 5736 pixels, that are 1 layer deep, and that we have the Interrupted Goode Homolosine  Coordinate reference system.  
Plot the blank Spatraster using the ggplot for the MS_area. Unlike the vector files which show empty counties with no values, the raster shows total values of each grid in the default color pallette. 

```{r}
ggplot() +
  geom_spatraster(data=MS_Area)

```

Like shapefiles, the ggplot interface using tidyterra can be used to change the colors to a custom pallette. I did this below using a tan to brown continum with the scale fill gradient option. For more information, type ?scale_fill_gradient into the console and learn about gradient colors. When you run the code, you will see that there is a huge variance in the scale, with soil carbon values being over 600 in the swamp values. These comprise very small values of the area, but the influence the color scheme and plot.

```{r}
ggplot() +
  geom_spatraster(data=MS_Area) +
  scale_fill_gradient(low = "tan", high = "brown")

```

Change the visual scale using the clamp variable. Everything over the value that you set will display at the value you specify. This code also adds the legend name.

```{r echo=T, results='hide'}
#We are setting 100 Mg C/ha as our upper display limit. Change this value up and down 
#and decide if you have a preference 
clamped_raster <- clamp(MS_Area, upper = 100, 
                        #values If FALSE values outside the clamping range become NA, 
                        #if TRUE, they get the extreme values. See what it looks like if  Values=False
                        values = TRUE)

ggplot() +
  #change the raster value to the clamped raster
  geom_spatraster(data=clamped_raster) +
  #keep the same color gradient
  scale_fill_gradient(low = "tan", high = "brown", 
                      #add the title to the legend as well
                      name = "Mg C/ha")

```

Add the county shapefile on top so we have some visual boundaries to look at, and add a title and some themes (described above) to the plot, and put the legend at the bottom of the plot. 

```{r}

#Create the SOC Map
SOC_Map<-ggplot() +
  #Use the clamped raster
  geom_spatraster(data=clamped_raster) +
  #keep the same color gradient
  scale_fill_gradient(low = "tan", high = "brown", 
                      #Add the units to the legend
                      name = "Mg C/ha", 
                      #make the NA values transparent, rather than grey for clarity
                      na.value = "transparent") +
  #Add the reprojected spatvector file using a fill value of NA, so you can see the underlying raster values
  geom_spatvector(data=ag_data1, fill=NA) + 
  #Add a title 
  ggtitle("Total Soil Organic Carbon") +
  #center the title and put the legend on the bottom. 
  theme(plot.title = element_text(hjust=0.5), legend.position = "bottom")

#check the map
SOC_Map

```

Use what we learned above to plot a side by side of the raster and vector data in the same figure

```{r}
#Move the crop_plot legend to the bottom so it matches the total SOC plot
crop_plot <- crop_plot + theme(legend.position="bottom")

#Use two upright columns to map the figures together. Try changing ncol to nrow to see the difference.
grid.arrange(crop_plot, SOC_Map, ncol=2)

```

## Your Turn

1)  Pull the soil clay from the top 5 cm map (in g/kg) from the geodata package: Hint, ?soil_world_vsi if you need more information.

2) Check the file Resolution and Projection, are they the same as the clamped_raster we used? What could the different resolutions mean?

3) Project the data to the existing dataset (if need be)

4) crop and mask the soil clay raster to the Mississippi river area

5) Customize and label your raster plot

6) plot it next to the SOC carbon map above
